Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-06-27 paper/main.tex
add reproducibility pipeline
make proof-read adjustments

In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}. The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive. Therefore, many techniques have been proposed for approximating $\phi$. For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smoothed spectral density

@@

\paragraph{New contributions} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the algorithms from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials, and realize that an efficient implementation of all of their methods can be summarized in a single algorithm. Based on the analysis of the randomized trace estimators for parameter-dependent matrices we provide bounds on the approximation error for our method.

@@

\paragraph{Reproducibility} This document, along with all included plots and tables, was generated from commit \href{https://github.com/FMatti/Rand-SD/tree/ca391a8}{ca391a8} at \href{https://github.com/FMatti/Rand-TRACE}{github.com/FMatti/Rand-TRACE} on 2024-02-05 at 18:44:29 UTC.

@@

The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Minkowski's integral inequality to extend the results to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
\begin{proof} 
    We define the 1-query Girard-Hutchinson estimate 
    \begin{equation}
        r(\mtx{B}(t), \vct{\psi}) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
    \end{equation}
    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.

@@

\begin{proof}
    From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
        \leq \lVert \mtx{A} \rVert _2^2 \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \vct{\omega}_1 \rVert _2^2 \right] + \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2^2 \right]
    \end{equation}
    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$. Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
    \begin{align}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
        &\leq \lVert \mtx{A} \rVert _2^2 (k + p) \left( \lVert \mtx{I}_k \rVert _2^2 + \frac{1}{k} \lVert \mtx{I}_k \rVert _F^2 \right) + (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right) \notag \\
        &\leq (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
    \end{align}

@@

The algorithm also accomodates the two other estimators introduced in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}, and the complexity for each evaluation of the estimate decreases significantly. For $n_{\mtx{\Psi}} = 0$ we end up with the Nyström estimator \refequ{equ:nystrom-trace-estimator}, provided the implementation evaluates $0/0$ to $0$ in the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}.

>>>>

In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}. The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$ with the Dirac delta distribution $\delta$. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive. Therefore, many techniques have been proposed for approximating $\phi$. For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smoothed spectral density

@@

\paragraph{New contributions} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude. Based on the analysis of the randomized trace estimators for parameter-dependent matrices we provide bounds on the approximation error for our method.

@@

\paragraph{Reproducibility} \input{re-pro-badge.tex}

@@

The idea behind the proof is to bound the higher order moments of the residual $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ for fixed values of $t$ and then use Minkowski's integral inequality to extend the results to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
\begin{proof} 
    We define the residual of the 1-query Girard-Hutchinson estimate 
    \begin{equation}
        r(\mtx{B}(t), \vct{\psi}) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
    \end{equation}
    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.

@@

\begin{proof}
    In the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms}, it is shown that Slepian's inequality \cite[theorem 7.2.1]{vershynin-2018-random-processes} applied to two appropriately constructed Gaussian fields yields
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
        \leq \lVert \mtx{A} \rVert _2^2 \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \vct{\omega}_1 \rVert _2^2 \right] + \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2^2 \right]
    \end{equation}
    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$. Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
    \begin{align}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
        &\leq \lVert \mtx{A} \rVert _2^2 (k + p) \left( \lVert \mtx{I}_k \rVert _2^2 + \frac{1}{k} \lVert \mtx{I}_k \rVert _F^2 \right) + (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right) \notag \\
        &\leq (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
    \end{align}

@@

The algorithm also accomodates the two other estimators introduced in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}, and the complexity for each evaluation of the estimate decreases significantly. For $n_{\mtx{\Psi}} = 0$ we end up with the Nyström estimator \refequ{equ:nystrom-trace-estimator}, provided the implementation evaluates $0/0$ to $0$ in the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}. Notice that in this case, \refalg{alg:nystrom-chebyshev-pp} is equivalent to the \enquote{spectrum sweeping} method from \cite[algorithm 5]{lin-2017-randomized-estimation}, but has a significantly lower computational complexity.
